Nix for Polyglot, Reproducible Data Science Workflows

### 2025-10-22 paper.qmd
Update paper.qmd

## Introduction: Reproducibility is also about software {#sec-intro}

@@

Once the image is built, the analysis can be executed inside a *container*,
which is a running instance of the image. Containers are typically run
non-interactively, allowing the complete computational environment to be
instantiated and executed with a single command.

@@

By leveraging these principles, [Nix]{.pkg} can effectively replace and integrate the functionalities of 
tools like [renv]{.pkg} and [Docker]{.pkg} for [R]{.proglang} projects, or `requirements.txt` files 
and virtual environments for [Python]{.proglang} projects. Furthermore, [Nix]{.pkg} excels at building
multi-language environments, which can seamlessly combine [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang},
a [LaTeX]{.proglang} distribution, or any of the numerous other tools available in `nixpkgs`. 
This enables the creation of a truly complete, project-specific, and deeply reproducible environment
that can be used interactively for development or non-interactively for automated analysis. 
As long as the `nixpkgs` repository (or a suitable fork, as discussed in Section @sec-fork) remains accessible,
the environment can be rebuilt reliably in the future.

@@

This expression can be saved in a file called `default.nix`. The environment can
then be built on a system with [Nix]{.pkg} installed using the `nix-build`
command.^[For installing [Nix]{.pkg}, we recommend the Determinate Systems
installer: \url{https://determinate.systems/posts/determinate-nix-installer}].
Once the build completes, the user can enter the interactive shell with
`nix-shell`. This shell contains all the packages specified in `default.nix` and
can be used for development, similar to activating a virtual environment in the
[Python]{.proglang} ecosystem.

@@

The package can then generate a visual representation of the pipeline's directed
acyclic graph:

@@

Figure @fig-dag can be generated by calling `rxp_ggdag()`.

@@

This appendix provides the complete source code for the polyglot pipeline
example discussed in Section 5. The pipeline simulates a Real Business Cycle
(RBC) model in [Julia]{.proglang}, trains an [XGBoost]{.pkg} model in
[Python]{.proglang}, and visualizes the results in [R]{.proglang}.

It is possible to run this complete example by executing the shell
script called `run_polyglot_example.sh` found in the repository of
this 
paper^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/run_polyglot_example.sh}].

#### The Environment Definition

This code snippet uses [rix]{.pkg} to generate a `default.nix` file that contains all
necessary dependencies for [R]{.proglang}, [Julia]{.proglang}, and
[Python]{.proglang}. It is typically saved in a file called `gen-env.R`:

```{=tex}
\begin{CodeInput}
# This script defines the polyglot environment our pipeline will run in.
library(rix)

# Define the complete execution environment
rix(
 # Pin the environment to a specific date for reproducibility.
 date = "2025-10-14",

 # R Packages
 r_pkgs = c(
   "ggplot2", "dplyr", "arrow", "rix", "rixpress", "quarto"
 ),

 # Julia Configuration
 jl_conf = list(
   jl_version = "lts",
   jl_pkgs = c("Distributions", "DataFrames", "Arrow", "Random")
 ),

 # Python Configuration
 py_conf = list(
   py_version = "3.13",
   py_pkgs = c("pandas", "scikit-learn", "xgboost", "pyarrow")
 ),

 project_path = ".",
 overwrite = TRUE
)
\end{CodeInput}
```

#### The RBC Model Simulation

This Julia script implements the state-space solution to the RBC model as a pure
function, and is saved in a subfolder of the project named `functions` and in a script
`functions.jl`. The economic theory is based on @depaoli2009rbc.

```{=tex}
\begin{Code}
# This script contains the helper functions for the Julia portion of the pipeline.
# It implements the correct, stable, and canonical log-linearized solution to
# a standard Real Business Cycle (RBC) model, based on the provided lecture slides.

# In an interactive session, you should uncomment the line below
# to load the required packages. rixpress handles this automatically.
#using LinearAlgebra, Distributions, DataFrames, Arrow, Random

# 1. Main Simulation Function

"""
    simulate_rbc_model(α, β, δ, ρ, σ, σ_z)

    Takes RBC model parameters as input, solves for the state-space representation
    using the method of undetermined coefficients (as per the slides), simulates
    the model for 250 quarters, and returns a DataFrame.
"""
function simulate_rbc_model(α, β, δ, ρ, σ, σ_z)

    # --- STEADY-STATE VALUES AND CONSTRUCTED PARAMETERS (Slides p. 16, 31) ---
    y_k = ((1/β) - 1 + δ) / α

    # --- SOLVE FOR POLICY FUNCTION COEFFICIENTS (Slides p. 19-22) ---
    # We solve for the coefficients of the policy functions:
    # c_hat_t = c_ck * k_hat_{t-1} + c_cz * z_hat_t
    # k_hat_t = c_kk * k_hat_{t-1} + c_kz * z_hat_t

    # Solve the quadratic equation for c_ck (coefficient of capital on consumption)
    # Based on Campbell (1994) and slide 22, we take the stable root.
    # a*x^2 + b*x + c = 0
    a_quad = y_k - δ
    b_quad = -( (1-β*(1-δ))*(y_k - δ) + y_k*(1+α) + 1 )
    c_quad = y_k * α

    # The stable solution is the smaller positive root
    c_ck = (-b_quad - sqrt(b_quad^2 - 4*a_quad*c_quad)) / (2*a_quad)

    # Now solve for the other coefficients based on c_ck
    c_kk = (y_k * α) / (y_k - δ - c_ck)
    c_cz = (y_k * (1 - c_kk * (1 - α) * β * ρ)) /
           ( (y_k - δ - c_ck) * (1 - β * ρ) + σ * (1 - c_ck) * (1 - β * ρ) )
    c_kz = (c_cz * (1 - c_ck)) / (y_k * (1 - α))

    # --- BUILD STATE-SPACE MATRICES FROM SOLVED COEFFICIENTS ---
    # State vector: s_t = [k_{t-1}, z_t]'
    # Transition:   s_t = T*s_{t-1} + R*ε_t

    T = [c_kk  c_kz * ρ
         0     ρ      ]

    R = [c_kz ; 1]

    # Observation Matrix for [output, consumption, investment]
    # y_hat_t = α*k_{t-1} + z_t
    # c_hat_t = c_ck*k_{t-1} + c_cz*z_t
    # i_hat_t = (i_y_ss)^-1 * (k_t - (1-δ)k_{t-1})
    #         = (i_y_ss)^-1 * ( (c_kk - (1-δ))k_{t-1} + c_kz*z_t )

    i_y_ss = δ * (α / ((1/β) - 1 + δ))

    C = [α      1
         c_ck   c_cz
         (c_kk - (1-δ))/i_y_ss   c_kz/i_y_ss]

    # --- SIMULATE THE MODEL ---
    Random.seed!(1234)
    n_periods = 250
    shocks = randn(n_periods) * σ_z

    states = zeros(2, n_periods)
    for t in 2:n_periods
        # The state is [k_{t-1}, z_t]'
        # To get the next state [k_t, z_{t+1}]', we first find k_t
        k_t = T[1,1]*states[1, t] + T[1,2]*states[2, t] + R[1]*shocks[t]
        z_tp1 = T[2,1]*states[1, t] + T[2,2]*states[2, t] + R[2]*shocks[t]

        states[1, t] = k_t # This is now k_t, which is "k_lag" for t+1
        states[2, t] = z_tp1 # This is now z_{t+1}
    end
    # Re-aligning states to be [k_{t-1}, z_t]
    k_lag = [0; states[1, 1:end-1]]
    z = [0; states[2, 1:end-1]]

    observables = C * [k_lag'; z']

    df = DataFrame(
        period = 1:n_periods,
        output = observables[1, :],
        consumption = observables[2, :],
        investment = observables[3, :],
        capital = k_lag,
        technology = z
    )

    return df
end

# 2. Encoder Function (Unchanged)
"""
    arrow_write(df::DataFrame, path::String)
    Encoder function to save a Julia DataFrame to an Arrow file.
"""
function arrow_write(df::DataFrame, path::String)
    Arrow.write(path, df)
end
\end{Code}
```

#### The XGBoost Forecasting Model

This Python script contains a series of modular functions to handle feature
engineering, data splitting, model training, and prediction, and is saved in
`functions/functions.py`.

```{=tex}
\begin{Code}
# This script contains modular helper functions for the Python portion of the pipeline.
# Each function performs a single, distinct task in the ML workflow.

# 1. Load required Python packages
# This is only needed if you run this script by hand,
# in which case, uncomment lines 8 to 11.
# With rixpress, the packages get loaded automatically.
#import pandas as pd
#import pyarrow.feather as feather
#from sklearn.model_selection import train_test_split
#import xgboost as xgb
# This script contains the helper functions for the Python portion of the pipeline.
# It defines the ML model training and prediction logic as a pure function.
# This script contains the helper functions for the Python portion of the pipeline.
# It defines the ML model training and prediction logic as a pure function.

# Step 1: Feature Engineering
def prepare_features(simulated_df: pd.DataFrame) -> pd.DataFrame:
    """Takes the raw simulated data and creates lagged features and the target variable."""
    df = simulated_df.copy()

    # Create lagged features
    for col in ['output', 'consumption', 'investment', 'capital', 'technology']:
        df[f'{col}_lag1'] = df[col].shift(1)

    # Drop the first row which now has NaNs
    df.dropna(inplace=True)

    return df

# Step 2: Data Splitting Functions
def get_X_train(processed_df: pd.DataFrame):
    """Gets the training features (X_train) from the processed data."""
    features = [col for col in processed_df.columns if '_lag1' in col]
    X = processed_df[features]
    train_size = int(0.75 * len(X))
    return X[:train_size]

def get_y_train(processed_df: pd.DataFrame):
    """Gets the training target (y_train) from the processed data."""
    y = processed_df['output']
    train_size = int(0.75 * len(y))
    return y[:train_size]

def get_X_test(processed_df: pd.DataFrame):
    """Gets the testing features (X_test) from the processed data."""
    features = [col for col in processed_df.columns if '_lag1' in col]
    X = processed_df[features]
    train_size = int(0.75 * len(X))
    return X[train_size:]

def get_y_test(processed_df: pd.DataFrame):
    """Gets the testing target (y_test) from the processed data."""
    y = processed_df['output']
    train_size = int(0.75 * len(y))
    return y[train_size:]

# Step 3: Model Training
def train_model(X_train: pd.DataFrame, y_train: pd.Series):
    """Initializes and trains the XGBoost model."""
    model = xgb.XGBRegressor(
        objective='reg:squarederror',
        n_estimators=100,
        learning_rate=0.1,
        max_depth=3,
        random_state=42
    )
    model.fit(X_train, y_train)
    return model

# Step 4: Prediction
def make_predictions(model, X_test: pd.DataFrame):
    """Makes predictions on the test set using the trained model."""
    return model.predict(X_test)

# Step 5: Format Final Results
def format_results(y_test: pd.Series, predictions) -> pd.DataFrame:
    """Combines test data and predictions into a final DataFrame for R."""
    results_df = pd.DataFrame({
        'period': y_test.index,
        'actual_output': y_test.values,
        'predicted_output': predictions
    })
    return results_df

# Encoder Function (for saving the final output)
def save_arrow(df: pd.DataFrame, path: str):
    """Encoder function to save a pandas DataFrame to an Arrow file."""
    feather.write_feather(df, path)

\end{Code}
```

### The Visualization

This R snippet uses [ggplot2]{.pkg} to create a plot comparing the model's output
against the XGBoost predictions and is saved in `functions/functions.R`.

```{=tex}
\begin{CodeInput}
# This script contains the helper functions for the R portion of the pipeline.
# It defines the visualization logic using ggplot2.

# 1. Load required R packages
# This is only needed if you run this script by hand,
# in which case, uncomment lines 8 and 9.
# With rixpress, the packages get loaded automatically.
#library(ggplot2)
#library(dplyr)

#-------------------------------------------------------------------------------
# 2. Main Function: Create the Visualization
#-------------------------------------------------------------------------------

#' Create a plot comparing actual vs. predicted output
#'
#' This function takes a data frame with actual and predicted time series data
#' and generates a ggplot visualization.
#'
#' @param predictions_df A data frame containing columns: 'period',
#'   'actual_output', and 'predicted_output'. This data frame will be the
#'   output of the Python XGBoost step.
#' @return A ggplot object.
#'
plot_predictions <- function(predictions_df) {
  # Create the plot object
  p <- ggplot(predictions_df, aes(x = period)) +

    # Add a line for the actual output from the RBC model simulation
    geom_line(
      aes(y = actual_output, color = "Actual (RBC Model)"),
      linewidth = 1
    ) +

    # Add a dashed line for the XGBoost model's predictions
    geom_line(
      aes(y = predicted_output, color = "Predicted (XGBoost)"),
      linetype = "dashed",
      linewidth = 1
    ) +

    # Define custom colors and legend labels
    scale_color_manual(
      name = "Series",
      values = c("Actual (RBC Model)" = "blue", "Predicted (XGBoost)" = "red")
    ) +

    # Add informative labels and a title
    labs(
      title = "XGBoost Prediction of RBC Model Output",
      subtitle = "Forecasting next-quarter output based on current-quarter economic variables",
      x = "Time (Quarters)",
      y = "Output (Log-deviations from steady state)"
    ) +

    # Use a clean theme
    theme_minimal() +

    # Improve legend position
    theme(legend.position = "bottom")

  # Return the final ggplot object
  return(p)
}
\end{CodeInput}
```

### Running the project

The structure of the project should look like this:

```bash
.
├── functions
│   ├── functions.jl
│   ├── functions.py
│   └── functions.R
├── gen-env.R
└── gen-pipeline.R

2 directories, 5 files
```

Assuming we are working on a system that has Nix installed, we can "drop" into
a temporary Nix shell with R and `{rix}` available:

```r
nix-shell --expr "$(curl -sl https://raw.githubusercontent.com/ropensci/rix/main/inst/extdata/default.nix)"
```

Once the shell is ready, start R (by simply typing `R`) and run `source("gen-env.R")`
to generate the adequate `default.nix` file. This is the environment that we can use
to work interactively with the code while developing, and in which the pipeline will
be executed. We can then leave the R session and the shell entirely (by typing `exit`)
and type `nix-shell` to drop into the environment as defined by the `default.nix`.
The pipeline can now be built using `source("gen-pipeline.R")`.

>>>>

## Introduction: Reproducibility is also about software {#secntro}

@@

Once the image is built, the analysis can be executed inside a *container*,
which is a running instance of the image. Containers are typically run
nonnteractively, allowing the complete computational environment to be
instantiated and executed with a single command.

@@

By leveraging these principles, [Nix]{.pkg} can effectively replace and integrate the functionalities of 
tools like [renv]{.pkg} and [Docker]{.pkg} for [R]{.proglang} projects, or `requirements.txt` files 
and virtual environments for [Python]{.proglang} projects. Furthermore, [Nix]{.pkg} excels at building
multi-language environments, which can seamlessly combine [R]{.proglang}, [Python]{.proglang}, [Julia]{.proglang},
a [LaTeX]{.proglang} distribution, or any of the numerous other tools available in `nixpkgs`. 
This enables the creation of a truly complete, project-specific, and deeply reproducible environment
that can be used interactively for development or nonnteractively for automated analysis. 
As long as the `nixpkgs` repository (or a suitable fork, as discussed in Section @sec-fork) remains accessible,
the environment can be rebuilt reliably in the future.

@@

This expression can be saved in a file called `default.nix`. The environment can
then be built on a system with [Nix]{.pkg} installed using the `nix-build`
command.^[For installing [Nix]{.pkg}, we recommend the Determinate Systems
installer: \url{https://determinate.systems/posts/determinate-nixnstaller}].
Once the build completes, the user can enter the interactive shell with
`nix-shell`. This shell contains all the packages specified in `default.nix` and
can be used for development, similar to activating a virtual environment in the
[Python]{.proglang} ecosystem.

@@

@@

The package can then generate a visual representation of the pipeline's directed
acyclic graph by running `rxp_dag()`, as can be seen in Figure @fig-dag.

@@

This appendix provides a conceptual walk-through of the polyglot pipeline example discussed in Section 5. 
The pipeline simulates a Real Business Cycle (RBC) model in [Julia]{.proglang}, 
trains an [XGBoost]{.pkg} model in [Python]{.proglang}, visualizes the results in [R]{.proglang}, 
and compiles a final report with [Quarto]{.pkg}.

The complete, runnable source code for this example is available in the paper's GitHub repository. 
A shell script, `run_polyglot_example.sh`, is provided to automate the entire process from environment 
creation to final output, as described at the end of this section^[\url{https://github.com/b-rodrigues/rix_paper/blob/master/run_polyglot_example.sh}].

#### Project Structure and Components

The project is organized into a set of scripts, each with a distinct responsibility. 
The core logic is separated from the environment definition and pipeline orchestration.

*   `functions/`: This directory contains the helper scripts with the core analytical code for each language.
    *   `functions.jl`: The [Julia]{.proglang} script for the economic simulation.
    *   `functions.py`: The [Python]{.proglang} script for the machine learning workflow.
    *   `functions.R`: The [R]{.proglang} script for data visualization.
*   `gen-env.R`: An [R]{.proglang} script that defines the reproducible computational environment.
*   `gen-pipeline.R`: The main [R]{.proglang} script that defines and orchestrates the entire multi-language pipeline.

#### Step 1: The Environment Definition

The foundation of the project is the reproducible environment, defined in `gen-env.R`.
This script uses the `rix()` function to programmatically generate a `default.nix` file. 
This file serves as a complete blueprint for the computational environment, specifying:

*   The exact versions of [R]{.proglang}, [Python]{.proglang}, and [Julia]{.proglang}.
*   A list of required packages for each language (e.g., `ggplot2` for R, `xgboost` for Python, `DataFrames` for Julia).
*   System-level dependencies like [Quarto]{.pkg}.

Crucially, the entire environment is pinned to a specific date (`2025-10-14`), 
ensuring that anyone who builds the environment, now or in the future, will get
the exact same software versions, guaranteeing reproducibility.

By dropping into a temporary shell using the following command:

it is possible to generate the `default.nix` by sourcing `gen-env.R`. Then, one
leaves this temporary shell, and builds the environment using the command `nix-shell`, 
which also drop the user into the development shell.

#### Step 2: The Core Analytical Logic

The scientific logic for each stage of the pipeline is encapsulated in the separate 
helper scripts within the `functions/` directory.

*   **The RBC Model Simulation (`functions.jl`):** This [Julia]{.proglang} script contains a single pure
  function that implements the state-space solution to the RBC model, based on the economic theory of 
  @depaoli2009rbc. It takes the model's economic parameters as inputs and returns the simulated time-series
  data as a DataFrame.
*   **The XGBoost Forecasting Model (`functions.py`):** This [Python]{.proglang} script handles the 
  complete machine learning workflow through a series of modular functions. Its responsibilities include
  feature engineering (creating lagged variables), splitting the data into training and testing sets, 
  training the [XGBoost]{.pkg} model, and generating predictions.
*   **The Visualization (`functions.R`):** This [R]{.proglang} script contains a function that uses
  `[ggplot2]{.pkg}` to create the final visualization. It is designed to take the data frame of 
  actual and predicted values produced by the Python script and generate a plot comparing the two series.

#### Step 3: Orchestrating the Pipeline

The entire workflow is defined and orchestrated by `gen-pipeline.R`. This script acts as the master plan, 
using functions from the `[rixpress]{.pkg}` package to define each computational step as a *derivation*.
It declaratively outlines the dependencies between steps:

1.  It begins by defining the RBC model parameters in [Julia]{.proglang}.
2.  It specifies that the **RBC simulation** in `functions.jl` depends on these parameters.
3.  It then defines the series of **Python steps** (feature preparation, training, prediction), 
  making each one dependent on the output of the previous one. The first Python step is explicitly 
  made dependent on the output from the Julia simulation. `rixpress` handles the passing of data 
  between languages, in this case using the Arrow file format for efficiency.
4.  Finally, it defines the **R visualization** step, which depends on the final predictions from 
  the Python model, and a **Quarto rendering** step that depends on the generated plot.

When this script is run in the development shell (by executing `source("gen-pipeline.R")`), 
[rixpress]{.pkg} translates the declared pipeline into a master Nix expression
that Nix can execute, automatically handling the caching of results and re-running only the necessary 
steps if a piece of code or data changes.

#### Running the Project

To execute this entire example, a user with Nix installed can simply run the `run_polyglot_example.sh` 
script available in the root of the paper's repository. This script automates the full process:

1.  It first executes `gen-env.R` inside a temporary Nix shell to build the `default.nix` file.
2.  It then executes `gen-pipeline.R` inside the newly defined environment. This triggers Nix to run the
  entire polyglot pipeline in the correct order.

Upon completion, the script will have generated all intermediate artifacts and the final `readme.html` 
report containing the visualization.
